Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-11-17 intro.tex
changes to chapter 1

The Large Hadron Collider (LHC) at CERN has been colliding
particles since 2009 with each successive run increasing the centre-of-mass
energy of the collision (7-13.6 TeV). With the experiment operating
over the course of decades, there is a vast quantity of
experimental data collected that has to be compared with
theoretical predictions. Currently, the theory predictions
are made with the Standard Model (SM) of
particle physics which describes the fundamental particles
of nature and their interactions.

The standard paradigm for comparing
theory and experiment is to simulate the particle collisions
from the initial collision all the way through
to the detection of the hundreds of particles produced in the collision.
General purpose Monte Carlo event generators are the de-facto
tool designed to generate these simulated events.
With the large number of scattering events collected at
the LHC already, and with even more expected at the High Luminosity LHC,
the rate of event generation is at risk of
being outpaced by experimental data collection.

The process of generating simulated events can be broadly
split up into three segments:
the highly energetic hard scattering process, 
the subsequent cascading of particles (parton shower),
and the formation of bound states detected (hadronisation).
While hadronisation is based on phenomological models,
the hard scattering and parton shower are derived
from the SM and are theoretically well-motivated.

The hard scattering process can be formulated as a probability
for particles to collide, represented by matrix elements.
These matrix elements are formally calculated order-by-order
in perturbation theory with each order leading
to increasingly difficult computations. In practice,
these matrix elements are provided in libraries interfaced
to the event generators and evaluation is largely automated for many
of the most important processes. However, the evaluation time
for these complex expressions becomes problematic
when generating the large samples required to compare with
experiments, especially when a significant portion of time
spent simulating an event is spent on the matrix element
evaluation.

The focus of this thesis will be on applying modern
machine learning methods to accelerate the evaluation
of matrix elements, hence speeding up the event
generation process in general. Machine learning algorithms
have become ubiquitous in many fields and their applicability
to a wide range of problems have made them a popular choice
in the high energy physics community as well. By combining the
well understood behaviour of matrix elements in specific
kinematic regions with powerful machine learning
algorithms, the construction of physics-inspired machine
learning models unlocks higher levels of performance
than would otherwise be achievable.

The structure of this thesis is as follows: in this chapter
I will recap the necessary concepts from the SM,
in particular quantum chromodynamics (QCD), the sector
which governs the behaviour of quarks and gluons. Furthermore,
the relationship between theory and experiment will be elaborated on
with the introduction of cross-sections and their relation
to matrix elements.

In Chapter~\ref{chapter:qcd}, I will discuss the challenges that arise
in fixed-order perturbative calculations and the current methods
that have been adopted in the community to deal with them. Some of
these methods also form the theoretical foundation which we
embed into the machine learning algorithm. This procedure is 
described in detail in Chapter~\ref{chapter:fame1} for electron-positron
annihilation for tree-level matrix element emulation.
A similar philosophy is applied in Chapter~\ref{chapter:fame2}
where next-to-leading order QCD k-factors are emulated for
the same processes.
These two chapters discuss in detail the procedure for
constructing emulators for electron-positron annihilation,
however, the more relevant processes for the LHC are in
proton-proton initiated collisions. The extension
to this scenario is detailed in Chapter~\ref{chapter:fame3}
where we also explore using the
emulator in a novel implementation to replace traditional
matrix element providers in the event generator SHERPA.
Finally, I will summarise and conclude the thesis in
Chapter~\ref{chapter:conclusion}.

\section{The Standard Model of particle physics}
    The Standard Model of particle physics (SM) is our current
    best working theory to describe all known elementary particles
    as well as three of the four fundamental forces. Developed
    predominantly in the latter half of the 20th century, it is
    one of the most well tested theories we have in science today.
    Some highlights include the highly precise predictions of the
    anomalous magnetic moment of the electron, agreeing with experimental
    measurements to more than 10 significant figures \cite{Aoyama:2017uqe},
    and the discovery of the Higgs boson in 2012 by the ATLAS \cite{ATLAS:2012yve}
    and CMS experiments \cite{CMS:2012qbp} at the LHC
    which was theorised decades prior.

    The SM is a gauge quantum field theory (QFT) where particles
    are described as excitations in quantum fields. It can be specified
    by the gauge group
    \begin{equation}\label{eqn:SM_gauge}
        \mathrm{SU}(3)_{c} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}\, ,
    \end{equation}
    where subscripts denote the charges of the gauge groups. The first gauge
    group with colour charge $c$ describes the interactions of the
    strong force within the theory of quantum chromodynamics (QCD), which we
    will elaborate more on in Section~\ref{sec:qcd}. The second
    and third gauge groups represent the electroweak sector of the SM with
    charges $L$ for left and $Y$ for hypercharge. Under electroweak
    spontaneous symmetry breaking (EWSB), this product becomes
    \begin{equation}\label{eqn:SM_SSB}
        \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y} \xrightarrow{\mathrm{EWSB}} \mathrm{U}(1)_{\mathrm{EM}} \, ,
    \end{equation}
    giving rise to the electromagnetic and weak forces we observe.
    The effect of gravity is considered to be negligible on the scales
    considered in the SM and so it is not described.
    Each of the three fundamental forces described by the SM is mediated
    by the exchange of a gauge boson. The gauge bosons mediating the strong
    and EM force are the massless gluon $g$ and massless photon $\gamma$,
    respectively. For the weak force the gauge bosons are the $W^{\pm}$
    and $Z^{0}$ bosons which attain a mass through the Higgs mechanism \cite{Englert:1964et,Higgs:1964pj,Guralnik:1964eu}
    during EWSB, which elucidates the Higgs boson $H$.

    The matter content of the SM consists of fermions which can be split
    into quarks and leptons. Quarks are massive and experience the strong,
    weak, and EM force. Leptons are defined by their lack of colour charge,
    meaning they do not experience the strong force. Leptons can be separated
    into charged leptons (electron $e$, muon $\mu$, tau $\tau$
    and their antiparticles) which experience the weak and EM force, and 
    neutrinos (electron neutrino $\nu_{e}$, muon neutrino $\nu_{\mu}$, tau
    neutrino $\nu_{\tau}$ and their antiparticles) which only experience the
    weak force. Neutrinos are massless in the SM but this is not a requirement
    of the model. In fact they have been observed to have mass \cite{Super-Kamiokande:1998kpq,SNO:2002tuh}.

    The particle content of the SM is summarised in Figure \ref{fig:SM_particles}
    showing the quarks, leptons and bosons along with their masses, charges
    and spin.
    \begin{figure}
        \includegraphics[width=\linewidth]{introduction/sm_particles.pdf}
        \caption{Particle content of the SM split into quarks, leptons,
        gauge bosons and scalar bosons. The columns for the fermions depict
        the three different generations. Masses, electric charge, and spin
        are given for each particle. The yellow contours indicate the coupling
        of bosons to fermions, illustrating the forces experienced by the
        fermions inside the contour. Figure reference \cite{SM_figure}.}
        \label{fig:SM_particles}
    \end{figure}
    The interactions of these fields are governed by the SM Lagrangian
    \footnote{technically Lagrangian density but we use the terms
    Lagrangian and Lagrangian density interchangeably}
    which can be written as
    \begin{equation}\label{eqn:L_SM}
        \mathcal{L}_{\mathrm{SM}} = \mathcal{L}_{\mathrm{gauge}} + \mathcal{L}_{\mathrm{fermion}} + \mathcal{L}_{\mathrm{Higgs}} + \mathcal{L}_{\mathrm{Yukawa}} + \mathcal{L}_{\mathrm{GF}} + \mathcal{L}_{\mathrm{ghost}} \, ,
    \end{equation}
    where $\mathcal{L}_{\mathrm{gauge}}$ describes the gauge fields,
    $\mathcal{L}_{\mathrm{fermion}}$ describes how fermions interact with
    gauge fields as well as their kinetic terms,
    $\mathcal{L}_{\mathrm{Higgs}}$ describes the Higgs field,
    $\mathcal{L}_{\mathrm{Yukawa}}$ describes the interaction between the Higgs
    field and fermions,
    $\mathcal{L}_{\mathrm{GF}}$ is a gauge fixing term,
    and $\mathcal{L}_{\mathrm{ghost}}$ is a ghost term.
    The last two terms are required to remove unphysical degrees of freedom
    when gauge fixing the theory.
    All terms in the SM Lagrangian are invariant under local transformations
    of the gauge group (\ref{eqn:SM_gauge}).

    In the following section we will focus on the gauge, fermion, gauge-fixing
    and ghost Lagrangian terms in the framework of QCD as that will be the
    most relevant sector for this thesis. The remaining terms are discussed at
    length in standard reference texts \cite{Peskin:1995ev,Schwartz:2014sze,Romao:2012pq}.

\section{Introduction to quantum chromodynamics}\label{sec:qcd}
    \subsection{The QCD Lagrangian}\label{sec:L_qcd}
    Quantum chromodynamics is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(N_{c} = 3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{s}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{s} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The generators are normalised by convention to be
    \begin{equation}\label{eqn:generator_normalisation}
        \mathrm{Tr}(T^{a}T^{b}) = \delta^{ab}T_{R} \, , \quad T_{R} = \dfrac{1}{2} \, .
    \end{equation}
    This normalisation sets the values of the Casimirs
    of the group as
    \begin{align}\label{eqn:casimirs}
        \begin{split}
            T^{a}_{ij}T^{a}_{jk} &= \delta_{ik}C_{F} \, , \quad C_{F} = \dfrac{N_{c}^2-1}{2N_{c}} \, , \\
            f^{abc}f^{abd} &= \delta^{cd}C_{A} \, , \quad C_{A} = N_{c} \, ,
        \end{split}
    \end{align}
    where $C_{F}=4/3$ and $C_{A}=3$ for QCD.
    They are collectively referred to as colour factors.

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in Section~\ref{sec:renormalisation}.
    The implication of this is that at collider
    experiments where collisions occur at extremely high energy,
    the strong coupling constant becomes small, a
    property known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$ (see Section~\ref{sec:xs}).

    To remove unphysical degrees of freedom from the
    theory we need to gauge-fix the theory
    and add a ghost Lagrangian. The gauge-fixing term
    in the $R_{\xi}$ gauges is written as
    \begin{equation}\label{eqn:L_GF}
        \mathcal{L}_{\mathrm{GF}} = -\dfrac{1}{2\xi}(\partial^{\mu}A_{\mu}^{a})^{2} \, ,
    \end{equation}
    where $\xi = 1$ corresponds to the Feynman-'t-Hooft gauge.
    The ghost Lagrangian is most commonly written
    in the Faddeev-Popov procedure as
    \begin{equation}\label{eqn:L_ghost}
        \mathcal{L}_{\mathrm{ghost}} = (\partial_{\mu}\bar{c}^{a})(\delta^{ac}\partial_{\mu} + g_{s}f^{abc}A^{b}_{\mu})c^{c} \, ,
    \end{equation}
    where $c^{a}$ ($\bar{c}^{a}$) are Faddeev-Popov ghosts (anti-ghosts).
    They are unphysical external states but must be
    included in internal lines for a Lorentz invariant
    perturbative gauge theory.

    \subsection{QCD Feynman rules}\label{sec:qcd_feynman}
    Now that we have written down the QCD Lagrangian,
    we can examine the terms to write down how the
    gauge bosons interact with the fermions, and especially
    important for QCD, the gauge boson self interactions. A convenient
    way to do this is by deriving the Feynman rules of the theory, which can be
    pieced together to form Feynman diagrams. We will see
    that Feynman diagrams are a simple but powerful tool
    to systematically build up all the possible ways a process
    can occur.

    Expanding (\ref{eqn:L_QCD}) and extracting the terms
    that mix the gauge and fermions fields, we get an interaction Lagrangian
    \begin{equation}\label{L_int}
        \mathcal{L}_{\mathrm{int}} = g_{s}A^{a}_{\mu}\bar{\psi}_{i}^{f}\gamma^{\mu}T^{a}_{ij}\psi_{j}^{f}-g_{s}f^{abc}(\partial_{\mu}A^{a}_{\nu})A^{b,\,\mu}A^{c,\,\nu} -\dfrac{g_{s}^{2}}{4}f^{abc}A^{b}_{\mu}A^{c}_{\nu}f^{ade}A^{d,\,\mu}A^{e,\,\nu} \, ,
    \end{equation}
    which we can interpret as follows: the first term
    is an interaction between a gluon, a quark, and an
    anti-quark, the second term is a triple gluon
    self-interaction, and the third term is a four gluon
    self-interaction. These mixing terms can be recast
    into Feynman rules representing interaction vertices.

    To connect vertices we require propagators, which
    we can read off from the kinetic Lagrangian where
    we collect terms from (\ref{eqn:L_QCD}) and (\ref{eqn:L_GF}), 
    \begin{equation}\label{L_kin}
        \mathcal{L}_{\mathrm{kin}} = -\dfrac{1}{4}(\partial_{\mu}A^{a}_{\nu}-\partial_{\nu}A^{a}_{\mu})^{2} - \dfrac{1}{2\xi}(\partial_{\mu}A^{a}_{\mu})^{2} + \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{\partial}-m_{f})\psi_{i}^{f} \, ,
    \end{equation}
    where the first two terms corresponds to a gluon
    propagator and the third term corresponds to
    a quark propagator.

    \begin{figure}
        \includegraphics[width=\linewidth]{introduction/qcd_feynman_rules.pdf}
        \caption{QCD Feynman rules with the Minkowski metric signature $g^{\mu\nu} = \mathrm{diag}(1, -1, -1, -1)$.}
        \label{fig:qcd_feynman_rules}
    \end{figure}

    From these Lagrangians, we can read off the QCD Feynman rules
    which we have collected in Figure~\ref{fig:qcd_feynman_rules}.
    Note that we have only written down the vertices and propagators
    for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

    From these Feynman rules, we identify that each
    quark-antiquark-gluon and triple gluon vertices
    are associated with one power of $\alpha_{s}$ (due to squaring $g_{s}$),
    and the four gluon vertex is associated with
    two powers of $\alpha_{s}$. This relationship
    between vertices and $\alpha_{s}$ allows
    us to systematically build up Feynman diagrams
    that correspond to a fixed order in $\alpha_{s}$.
    This point will be elaborated on in Section
    \ref{sec:xs} where we discuss scattering amplitudes
    and how they are related to Feynman diagrams.

>>>>

The Large Hadron Collider (LHC) at CERN has been colliding
particles since 2009, with each successive run increasing the centre-of-mass
energy of the collision (7-13.6 TeV). As the experiment has
operated for over a decade, there is a vast quantity of
experimental data collected that has to be compared with
theoretical predictions. Currently, the theory predictions
are made with the Standard Model (SM) of
particle physics which describes the fundamental particles
of nature and their interactions.

The standard paradigm for comparing
theory and experiment is to simulate the particle collisions
from the initial collision all the way through
to the detection of the hundreds of particles produced in the collision.
General purpose Monte Carlo event generators are the de facto
tool designed to generate these simulated events.
With the large number of scattering events collected at
the LHC already, and with even more expected at the High Luminosity LHC (HL-LHC),
the rate of event generation is at risk of
being outpaced by experimental data collection.

The process of generating simulated events can be broadly
split up into three segments:
the highly energetic hard scattering process, 
the subsequent cascading of particles (parton shower),
and the formation of bound states detected (hadronisation),
which makes their way to the detectors.
While hadronisation is based on phenomenological models,
the hard scattering and parton shower are derived
from the SM and are theoretically well-motivated.

The hard scattering process can be formulated as a probability
for particles to collide, represented by matrix elements.
These matrix elements are formally calculated order-by-order
in perturbation theory, with each order leading
to increasingly difficult computations at each order. In practice,
these matrix elements are provided in libraries interfaced
to the event generators and evaluation is largely automated for many
of the most important processes. However, the evaluation time
for these complex expressions becomes problematic
when generating the large samples required to compare with
experiments. Especially when a significant portion of time
spent simulating an event is spent on evaluating matrix
elements.

The focus of this thesis will be on applying modern
machine learning methods to accelerate the evaluation
of matrix elements in order to speed up the event
generation process. Machine learning algorithms
have become ubiquitous in many fields, and their applicability
to a wide range of problems have made them a popular choice
in the high energy physics community as well. By combining the
well understood behaviour of matrix elements in specific
kinematic regions with powerful machine learning
algorithms, the construction of physics-inspired machine
learning models unlocks higher levels of performance
than would otherwise be achievable.

The structure of this thesis is as follows: in this chapter
I will recap the necessary concepts from the SM,
in particular quantum chromodynamics (QCD), the sector
which governs the behaviour of quarks and gluons. Furthermore,
I will elaborate on the relationship between theory and experiment
with the introduction of cross-sections, and their relation
to matrix elements.
In Chapter~\ref{chapter:qcd}, I will discuss the challenges that arise
in fixed-order perturbative calculations and the current methods
that have been adopted in the community to deal with them.
I will briefly introduce Monte Carlo event generators in
Chapter~\ref{chapter:MC}, where the theory discussed in the first
two chapters is applied in practice. In Chapter~\ref{chapter:ml},
I discuss how bottlenecks in traditional event generator techniques
motivates the usage of machine learning based approaches, with
particular attention placed on using neural networks as emulators
for matrix elements.
This construction of this neural network emulator is
described in detail in Chapter~\ref{chapter:fame1} for tree-level
electron-positron annihilation matrix element emulation.
A similar philosophy is applied in Chapter~\ref{chapter:fame2}
where next-to-leading order QCD K-factors are emulated for
the same processes.
These two chapters discuss in detail the procedure for
constructing emulators for electron-positron annihilation,
however, the more relevant processes for the LHC are in
proton-proton initiated collisions. The extension
to this scenario is detailed in Chapter~\ref{chapter:fame3}
where I also explore using the
emulator in a novel implementation to supplement traditional
matrix element providers in the event generator {\Sherpa}.
Finally, I will summarise and conclude the thesis in
Chapter~\ref{chapter:conclusion}.

\section{The Standard Model of particle physics}
    The Standard Model of particle physics is our current
    best working theory to describe all known elementary particles,
    as well as three of the four fundamental forces. Developed
    predominantly in the latter half of the 20th century, it is
    one of the most well tested theories that we have in science today.
    Some highlights include the highly precise predictions of the
    anomalous magnetic moment of the electron, which agrees with experimental
    measurements to more than 10 significant figures \cite{Aoyama:2017uqe},
    and the discovery of the Higgs boson in 2012 by the ATLAS \cite{ATLAS:2012yve}
    and CMS experiments \cite{CMS:2012qbp} at the LHC,
    which was theorised decades prior.

    The SM is a gauge quantum field theory (QFT) where particles
    are described as excitations of quantum fields. The symmetry group
    of the SM is
    \begin{equation}\label{eqn:SM_gauge}
        \mathrm{SU}(3)_{c} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}\, ,
    \end{equation}
    where subscripts denote the charges of the gauge groups. The first gauge
    group with colour charge $c$ describes the interactions of the
    strong force within the theory of quantum chromodynamics, which we
    will elaborate more on in Section~\ref{sec:qcd}. The second
    and third gauge groups represent the electroweak sector of the SM, where
    $L$ denotes left-chiral fields that carry weak isospin, whilst $Y$ denotes
    the weak hypercharge. Under electroweak
    spontaneous symmetry breaking (EWSB), this product becomes
    \begin{equation}\label{eqn:SM_SSB}
        \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y} \xrightarrow{\mathrm{EWSB}} \mathrm{U}(1)_{\mathrm{EM}} \, ,
    \end{equation}
    giving rise to the electromagnetic and weak forces that we observe.
    The effect of gravity is considered to be negligible on the scales
    considered in the SM, and so it is not described.
    Each of the three fundamental forces described by the SM is mediated
    by the exchange of a gauge boson. The massless gauge bosons mediating the strong
    and EM force are the gluon $g$ and photon $\gamma$,
    respectively. For the weak force the gauge bosons are the $W^{\pm}$
    and $Z^{0}$ bosons, which attain a mass through the Higgs mechanism \cite{Englert:1964et,Higgs:1964pj,Guralnik:1964eu}
    during EWSB, elucidating the Higgs boson $H$.

    The matter content of the SM consists of fermions, which can be split
    into quarks and leptons. Quarks are massive and experience the strong,
    weak, and EM force. Leptons are defined by their lack of colour charge,
    meaning they do not experience the strong force. Leptons can be separated
    into charged leptons (electron $e$, muon $\mu$, tau $\tau$
    and their antiparticles) which experience the weak and EM force, and 
    neutrinos (electron neutrino $\nu_{e}$, muon neutrino $\nu_{\mu}$, tau
    neutrino $\nu_{\tau}$ and their antiparticles) which only experience the
    weak force. Neutrinos are massless in the SM, however, they have been
    observed to have mass \cite{Super-Kamiokande:1998kpq,SNO:2002tuh}. This
    prompts physics beyond the Standard Model (BSM) to describe these observed masses.

    The particle content of the SM is summarised in Figure \ref{fig:SM_particles},
    which shows the quarks, leptons and bosons along with their masses, charges
    and spin.
    \begin{figure}
        \includegraphics[width=\linewidth]{introduction/sm_particles.pdf}
        \caption{Particle content of the SM, split into quarks, leptons,
        gauge bosons and scalar bosons. The columns for the fermions depict
        the three different generations. The masses, electric charge, and spin
        are given for each particle. The yellow contours indicate the coupling
        of bosons to fermions, illustrating the forces experienced by the
        fermions inside the contour. Figure reproduced from \cite{SM_figure}.}
        \label{fig:SM_particles}
    \end{figure}
    The interactions of these fields are governed by the SM Lagrangian\footnote{Technically Lagrangian density but we use the terms
    Lagrangian and Lagrangian density interchangeably.}
    which can be written as
    \begin{equation}\label{eqn:L_SM}
        \mathcal{L}_{\mathrm{SM}} = \mathcal{L}_{\mathrm{gauge}} + \mathcal{L}_{\mathrm{fermion}} + \mathcal{L}_{\mathrm{Higgs}} + \mathcal{L}_{\mathrm{Yukawa}} + \mathcal{L}_{\mathrm{GF}} + \mathcal{L}_{\mathrm{ghost}} \, ,
    \end{equation}
    where $\mathcal{L}_{\mathrm{gauge}}$ describes the gauge fields,
    $\mathcal{L}_{\mathrm{fermion}}$ describes how fermions interact with
    gauge fields as well as their kinetic terms,
    $\mathcal{L}_{\mathrm{Higgs}}$ describes the Higgs field,
    $\mathcal{L}_{\mathrm{Yukawa}}$ describes the interaction between the Higgs
    field and fermions,
    $\mathcal{L}_{\mathrm{GF}}$ is a gauge fixing term,
    and $\mathcal{L}_{\mathrm{ghost}}$ is a ghost term.
    The last two terms are required to remove unphysical degrees of freedom
    when gauge fixing the theory.
    All terms in the SM Lagrangian are invariant under local transformations
    of the gauge group Eq. (\ref{eqn:SM_gauge}).

    In the following section we will focus on the gauge, fermion, gauge-fixing
    and ghost Lagrangian terms in the framework of QCD, which is the most
    relevant sector for this thesis. The remaining terms are discussed at
    length in standard reference texts \cite{Peskin:1995ev,Schwartz:2014sze,Romao:2012pq}.

\section{Introduction to quantum chromodynamics}\label{sec:qcd}
    \subsection{The QCD Lagrangian}\label{sec:L_qcd}
    Quantum chromodynamics is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(N_{c} = 3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where repeated indices are summed over.
    The fields $\psi_{i}^{f}$ are the fermions field operators,
    representing quarks and antiquarks with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. The field operators transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{s}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{s} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation such that
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in the fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    By convention, the generators are normalised to be
    \begin{equation}\label{eqn:generator_normalisation}
        \mathrm{Tr}(T^{a}T^{b}) = \delta^{ab}T_{R} \, , \quad \mathrm{where} \quad T_{R} = \dfrac{1}{2} \, .
    \end{equation}
    This normalisation sets the values of the Casimirs
    of the group as
    \begin{align}\label{eqn:casimirs}
        \begin{split}
            T^{a}_{ij}T^{a}_{jk} &= \delta_{ik}C_{F} \, , \quad C_{F} = \dfrac{N_{c}^2-1}{2N_{c}} \, , \\
            f^{abc}f^{abd} &= \delta^{cd}C_{A} \, , \quad C_{A} = N_{c} \, ,
        \end{split}
    \end{align}
    where $C_{F}=4/3$ and $C_{A}=3$ for QCD.
    These are collectively referred to as colour factors.

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant is in fact not constant, and
    depends on the energy scale of the process. This
    is due to the process of renormalisation, which will be discussed
    in Section~\ref{sec:renormalisation}.
    The implication of this is that at collider
    experiments where collisions occur at extremely high energy,
    the strong coupling constant becomes small, a
    property known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$ (see Section~\ref{sec:xs}).
    This is normally precluded as $\alpha_{s} > 1$ at low
    energies.

    To remove unphysical degrees of freedom from the
    theory we need to fix the gauge
    and add a ghost Lagrangian. The gauge-fixing term
    in the $R_{\xi}$ gauge is written as
    \begin{equation}\label{eqn:L_GF}
        \mathcal{L}_{\mathrm{GF}} = -\dfrac{1}{2\xi}(\partial^{\mu}A_{\mu}^{a})^{2} \, ,
    \end{equation}
    where $\xi = 1$ corresponds to the Feynman-'t-Hooft gauge.
    The ghost Lagrangian is most commonly written
    in the Faddeev-Popov procedure as
    \begin{equation}\label{eqn:L_ghost}
        \mathcal{L}_{\mathrm{ghost}} = (\partial_{\mu}\bar{c}^{a})(\delta^{ac}\partial_{\mu} + g_{s}f^{abc}A^{b}_{\mu})c^{c} \, ,
    \end{equation}
    where $c^{a}$ ($\bar{c}^{a}$) are Faddeev-Popov ghosts (antighosts).
    The ghosts are unphysical external states but must be
    included in internal lines for a gauge invariant theory.

    \subsection{QCD Feynman rules}\label{sec:qcd_feynman}
    Now that we have written down the QCD Lagrangian,
    we can examine the terms that govern the interactions between
    bosons and fermions, and especially
    important for QCD, the gauge boson self interactions. A convenient
    way to do this is by deriving the Feynman rules of the theory, which can be
    pieced together to form Feynman diagrams. We will see
    that Feynman diagrams are a simple but powerful tool
    to systematically build up all the possible ways in which a process
    can occur.

    Expanding Eq. (\ref{eqn:L_QCD}) and extracting the terms
    that mix the gauge and fermions fields, we get an interaction Lagrangian
    \begin{equation}\label{L_int}
        \mathcal{L}_{\mathrm{int}} = g_{s}A^{a}_{\mu}\bar{\psi}_{i}^{f}\gamma^{\mu}T^{a}_{ij}\psi_{j}^{f}-g_{s}f^{abc}(\partial_{\mu}A^{a}_{\nu})A^{b,\,\mu}A^{c,\,\nu} -\dfrac{g_{s}^{2}}{4}f^{abc}A^{b}_{\mu}A^{c}_{\nu}f^{ade}A^{d,\,\mu}A^{e,\,\nu} \, ,
    \end{equation}
    which we can interpret as follows: the first term
    is an interaction between a gluon and two fermion field operators,
    the second term is a three gluon
    self-interaction, and the third term is a four gluon
    self-interaction. These mixing terms can be recast
    into Feynman rules representing interaction vertices.

    To connect vertices we require propagators, which
    we can read off from the kinetic Lagrangian where
    we collect terms from Eqns. (\ref{eqn:L_QCD}) and (\ref{eqn:L_GF}), 
    \begin{equation}\label{L_kin}
        \mathcal{L}_{\mathrm{kin}} = -\dfrac{1}{4}(\partial_{\mu}A^{a}_{\nu}-\partial_{\nu}A^{a}_{\mu})^{2} - \dfrac{1}{2\xi}(\partial_{\mu}A^{a}_{\mu})^{2} + \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{\partial}-m_{f})\psi_{i}^{f} \, ,
    \end{equation}
    where the first two terms corresponds to a gluon
    propagator and the third term corresponds to
    a quark propagator.

    \begin{figure}
        \includegraphics[width=\linewidth]{introduction/qcd_feynman_rules.pdf}
        \caption{QCD Feynman rules with the Minkowski metric $g^{\mu\nu} = \mathrm{diag}(1, -1, -1, -1)$.}
        \label{fig:qcd_feynman_rules}
    \end{figure}

    From these Lagrangians, we can read off the QCD Feynman rules
    which we have collected in Figure~\ref{fig:qcd_feynman_rules}.
    Note that we have only written down the vertices and propagators
    for the physical gluons and quarks.
    Ghosts also have associated Feynman rules but we do
    not specify those here. A full list of Feynman rules
    in the SM can be found in Ref. \cite{Romao:2012pq}.

    From these Feynman rules, we identify that each
    fermion-gluon and three gluon vertices
    are associated with one power of $\alpha_{s}$ (due to squaring $g_{s}$),
    and the four gluon vertex is associated with
    two powers of $\alpha_{s}$. This relationship
    between vertices and $\alpha_{s}$ allows
    us to systematically build up Feynman diagrams
    that correspond to a fixed-order in $\alpha_{s}$.
    This point will be elaborated on in Section
    \ref{sec:xs} where we discuss scattering amplitudes
    and how they are related to Feynman diagrams.
